The evolution of gene expression in the term placenta of viviparous mammals

### 2016-08-12 manuscript/manuscript.Rnw
add main mentioned genes table

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    alignment.species <- species.name
    if (alignment.species=="pan paniscus") {
        alignment.species <- "pan troglodytes"
    } else if (alignment.species=="ateles fusciceps") {
        alignment.species <- "homo sapiens"
    } else if (grepl("spalax",alignment.species)) {
        alignment.species <- "mus musculus"

@@

    temp <- 
        combined.fpkm[species==species.name &
                      !(!is.na(human_name) & 
                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
                             gene_short_name)
                      ][order(-mean_fpkm)][,list(name_or_id,

@@

                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
    temp <- temp[!duplicated(name_or_id)][1:n]
    temp[,used_alignment:=FALSE]
    temp[is.na(human_name) & !is.na(human_alignment_symbol),

@@

    temp[is.na(human_name),human_name:=human_alignment_symbol]
    setkey(temp,"human_name")
    temp <- 
        hugo.nomenclature[temp][order(-mean_fpkm),
                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,

@@

    temp[!is.na(name_or_id),

@@

    temp[!is.na(`Approved Name`),
         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
    temp[!is.na(`Approved Symbol`),
         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
    temp[,used_alignment:=NULL]
    temp[,sem_fpkm:=as.numeric(NA)]
    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
    temp[,sd_fpkm:=NULL]
    temp[,n_fpkm:=NULL]
    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
    if (temp[,all(is.na(`SEM`))]) {
        temp[,`SEM`:=NULL]
    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
        temp[,`Human Symbol`:=NULL]
    temp[,`FPKM`:=round(`FPKM`)]
    return(temp)
gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
    species.name <- gsub("_"," ",species.name)
    fpkm <- combined.fpkm[species==species.name &
                              (gene_short_name==gene | 
                               gene_id==gene),
                          mean_fpkm][1]
    if (is.na(fpkm)) {
        stop("FPKM is NA!!!!")
    if (units || complete)
        fpkm <- formatC(fpkm,digits=0,format="f")
    if (units) {
        fpkm <- paste0(fpkm,
                       "\\ \\text{FPKM}")
    if (complete) {
        if (any(grepl("^ENS.+G\\d+$",gene))) {
            fpkm <- paste0("\\text{",gene,"} = ",
                           fpkm)
        } else {
            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
                           fpkm)
    if (units || complete) {
        fpkm <- paste0("$",fpkm,"$")
    return(fpkm)

        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     align=c("c","c","p{2in}","r","r","r")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

@@

@@

>>>>

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    alignment.species <- species.name
    if (alignment.species=="pan paniscus") {
        alignment.species <- "pan troglodytes"
    } else if (alignment.species=="ateles fusciceps") {
        alignment.species <- "homo sapiens"
    } else if (grepl("spalax",alignment.species)) {
        alignment.species <- "mus musculus"
    }
    temp <- 
        combined.fpkm[species==species.name &
                      !(!is.na(human_name) & 
                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
                             gene_short_name)
                      ][order(-mean_fpkm)][,list(name_or_id,
                                                 human_alignment_symbol,
                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
    temp <- temp[!duplicated(name_or_id)][1:n]
    temp[,used_alignment:=FALSE]
    temp[is.na(human_name) & !is.na(human_alignment_symbol),
                    used_alignment:=TRUE]
    temp[is.na(human_name),human_name:=human_alignment_symbol]
    setkey(temp,"human_name")
    temp <- 
        hugo.nomenclature[temp][order(-mean_fpkm),
                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,
                                     mean_fpkm,sd_fpkm,n_fpkm)]
    temp[!is.na(name_or_id),
         name_or_id:=paste0("\\href{http://www.ensembl.org/",
                             capfirst(gsub(" ","_",alignment.species)),
                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
                             xtable.sanitize.text(name_or_id),"}}")]
    temp[!is.na(`Approved Name`),
         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
    temp[!is.na(`Approved Symbol`),
         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
    temp[,used_alignment:=NULL]
    temp[,sem_fpkm:=as.numeric(NA)]
    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
    temp[,sd_fpkm:=NULL]
    temp[,n_fpkm:=NULL]
    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
    if (temp[,all(is.na(`SEM`))]) {
        temp[,`SEM`:=NULL]
    }
    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
        temp[,`Human Symbol`:=NULL]
    }
    temp[,`FPKM`:=round(`FPKM`)]
    return(temp)
}
@

<<gene.expression,results="hide",echo=FALSE>>=
mentioned.genes <- load_cache(label="save_mentioned_genes",object="saved_mentioned_genes",list())
gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
    species.name <- gsub("_"," ",species.name)
    mentioned.genes[[paste0(species.name,"_",gene)]] <<-
        list("species.name"=species.name,
             "gene"=gene)
    fpkm <- combined.fpkm[species==species.name &
                              (gene_short_name==gene | 
                               gene_id==gene),
                          mean_fpkm][1]
    if (is.na(fpkm)) {
        stop("FPKM is NA!!!!")
    }
    if (units || complete)
        fpkm <- formatC(fpkm,digits=0,format="f")
    if (units) {
        fpkm <- paste0(fpkm,
                       "\\ \\text{FPKM}")
    }
    if (complete) {
        if (any(grepl("^ENS.+G\\d+$",gene))) {
            fpkm <- paste0("\\text{",gene,"} = ",
                           fpkm)
        } else {
            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
                           fpkm)
        }
    }
    if (units || complete) {
        fpkm <- paste0("$",fpkm,"$")
    }
    return(fpkm)
}

        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     align=c("c","p{0.5in}","p{1in}","r","r","r")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

<<mentioned.genes.table.functions,echo=FALSE,results="asis">>=
mentioned.genes.table <- list()
if (length(mentioned.genes)==0) {
    mentioned.genes.table[["No genes"]] <-
        data.table(species="No genes",
                   name_or_id="No genes",
                   human_alignment_symbol="NOGENE",
                   human_name="NONAME",
                   mean_fpkm = 0,
                   sd_fpkm = 1,
                   n_fpkm = 2)

@@

for (entry in names(mentioned.genes)) {
    mentioned.genes.table[[entry]] <-
        combined.fpkm[species==mentioned.genes[[entry]][["species.name"]] &
                      (gene_short_name==mentioned.genes[[entry]][["gene"]] | 
                       gene_id==mentioned.genes[[entry]][["gene"]]),
                      list(species,
                           name_or_id,

@@

                           human_name,
                           mean_fpkm,
                           sd_fpkm,
                           n_fpkm)]
mentioned.genes.table <-
    rbindlist(mentioned.genes.table)
mentioned.genes.table[,used_alignment:=FALSE]
mentioned.genes.table[is.na(human_name) & !is.na(human_alignment_symbol),

@@

mentioned.genes.table[is.na(human_name),human_name:=human_alignment_symbol]
setkey(mentioned.genes.table,"human_name")
mentioned.genes.table <- 
    hugo.nomenclature[mentioned.genes.table][,
                                             list(species,
                                                  name_or_id,
                                                  used_alignment,
                                                  `Approved Symbol`,
                                                  `Approved Name`,

@@

mentioned.genes.table[,alignment.species:=species]
mentioned.genes.table[alignment.species=="pan paniscus",alignment.species:="pan troglodytes"]
mentioned.genes.table[alignment.species=="ateles fusciceps",alignment.species:="homo sapiens"]
mentioned.genes.table[grepl("spalax",alignment.species),alignment.species:="mus musculus"]
mentioned.genes.table[!is.na(name_or_id),

@@

mentioned.genes.table[!is.na(`Approved Name`),
                      `Approved Name`:=paste0("\\href{http://www.ensembl.org/",
                                              "Homo_sapiens/Gene/Summary?g=",
                                              `Approved Symbol`,"}{",
                                              xtable.sanitize.text(`Approved Name`),
                                              "}")]
mentioned.genes.table[!is.na(`Approved Symbol`),
                      `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/",
                                                "Homo_sapiens/Gene/Summary?g=",
                                                `Approved Symbol`,
                                                "}{\\textit{",
                                                xtable.sanitize.text(`Approved Symbol`),
                                                "}}")]
mentioned.genes.table[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
mentioned.genes.table[,sem_fpkm:=as.numeric(NA)]
mentioned.genes.table[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
mentioned.genes.table[,used_alignment:=NULL]
mentioned.genes.table[,alignment.species:=NULL]
mentioned.genes.table[,sd_fpkm:=NULL]
mentioned.genes.table[,n_fpkm:=NULL]
mentioned.genes.table[,`Approved Name`:=NULL]
### order by species and then by FPKM
setorder(mentioned.genes.table,`Approved Symbol`,species,-mean_fpkm)
### italicize species names
mentioned.genes.table[,species:=paste0("\\textit{",gsub("^(.)","\\U\\1",species,perl=TRUE),"}")]
setnames(mentioned.genes.table,c("Species","Gene Symbol",
                                 "Human Symbol",
                                 ## "Human Name",
                                 "FPKM","SEM"))
mentioned.genes.table[,`FPKM`:=round(`FPKM`)]

output.mentioned.genes.table <- function(subset.table=quote(TRUE),
                                         human.symbol=TRUE) {
    mgt.subset <- mentioned.genes.table[eval(subset.table)]
    mgt.display <- c("s","s","s","s","d","d")
    mgt.align <- c("c","c",
                   ">{\\raggedright \\centering}p{0.5in}",
                   ">{\\raggedright \\centering}p{0.5in}",
                   ##">{\\raggedright}p{4in}",
                   "r","r")
    if (!human.symbol) {
        mgt.subset[,`Human Symbol`:=NULL]
        mgt.display <- mgt.display[-4]
        mgt.align <- mgt.align[-4]
    }

    print(xtable(mgt.subset,
                 display=mgt.display,
                 align=mgt.align
                 ),
          math.style.exponents=TRUE,
          booktabs=TRUE,
          floating=FALSE,
          sanitize.text.function=genes.table.sanitize,
          include.rownames=FALSE)
}
@

@@

\begin{table}
\footnotesize
<<primate.genes,echo=FALSE,results="asis">>=
output.mentioned.genes.table(subset.table=quote(grepl("Homo|Ateles|Pan",Species)),
                             human.symbol=FALSE
                             )
@ 
\caption{Notable genes ordered by species and \ac{FPKM}. Gene
  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
  name corresponding to the human symbol. A complete table of 
  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
\label{tab:notable_genes}
\end{table}

@@

<<save_mentioned_genes,echo=FALSE,results="hide">>=
saved_mentioned_genes <- mentioned.genes
@
